---
format: html
title: "Joining strings with missing data together in R"
author: Ben Harrap
date: 2025-09-24
toc: false
categories: 
  - R
  - programming
  - strings
---

Joining multiple string columns together in R isn't a particularly tricky thing to do. Until recently though, I found it unreasonably annoying trying to join strings where missing data is present. For example, let's say we've got the following address data:

```{r}
#| message: false
library(tidyverse)
addresses <- tibble(
  unit = c("Shop 4","Unit 185", NA, NA),
  street = c("2 Frencham Pl","20 Allara St","1 McCoy Cct","130 Garran Rd Acton ACT 2601"),
  suburb = c("Downer","Canberra","Acton", NA),
  postcode = c("2602","2601","2601", NA)
)
```

The first two have a value for each element, while the third address doesn't have a `unit`, and the fourth for some reason has all the elements in `street`. What I want to do is combine all of these address elements into a single string and ignore missing values. Let's review the ways we can go about this.

## `tidyr::unite()`

I'll cut to the chase – `tidyr::unite()` is perfect for this:

```{r}
addresses |>
  bind_cols(
    unite(
      data = addresses,
      col = "address",
      sep = " ",
      na.rm = TRUE
    )
  )
```

I only recently learned about `unite()` (thanks Stephen!) and was surprised to find it in `tidyr` and not in `stringr`. Anyway, it's perfect!

## `base::paste()`

We could use `paste()`, except that it includes `NA` as the letters NA in the output, which is not what I want.

```{r}
addresses |> 
  mutate(
    address = paste(
      unit, street, suburb, postcode,
      sep = " "
      )
  )
```

We could do some cleaning to overcome this issue:

```{r}
addresses |>
  replace_na(
    replace = list(unit = "", suburb = "", postcode = "")
  ) |>
  mutate(
    address = paste(
      unit, street, suburb, postcode,
      sep = " "
    )
  )
```

But the empty strings are now causing excessive separators to be included. We can fix this with `str_trim()` and `str_squish()`, but it's getting silly at this point. Just use `unite()`!

```{r}
addresses |>
  replace_na(
    replace = list(unit = "", suburb = "", postcode = "")
  ) |>
  mutate(
    address = str_trim(
      paste(
        unit, street, suburb, postcode,
        sep = " "
      )
    )
  )
```

## `base::sprintf()`

Another base option is to use `sprintf()`, which seems like we'd still need to do some extra work to handle the `NA`.

```{r}
addresses |> 
  mutate(
    address = sprintf(
      "%s %s %s %s",
      unit, street, suburb, postcode
      )
  )
```

## `stringr::str_c()`

We could use `str_c()` but we'll have the same issues as we did with `paste`. Notice though that `str_c` returns `NA` if any of the input strings are `NA`.

```{r}
addresses |> 
  mutate(
    address = str_c(
      unit, street, suburb, postcode,
      sep = " "
      )
  )
```

## `stringr::str_flatten()`

Interestingly, we can achieve the same result as `unite()` if we perform `str_flatten()` on a row-wise basis. I doubt this is a good idea though, as row-wise stuff tends to get very expensive, very quickly.

```{r}
addresses |> 
  rowwise() |> 
  mutate(
    address = str_flatten(
      c(unit, street, suburb, postcode),
      collapse = " ",
      na.rm = TRUE
      )
  )
```

## `glue::glue()`

`glue` is another option and arguably works better than `paste` or `str_c` because of the `.na` option letting us handle the `NA` within the function itself.

```{r}
library(glue)
addresses |> 
  mutate(
    address = glue(
      "{unit}","{street}","{suburb}","{postcode}",
      .na = "",
      .sep = " "
      )
  )
```

Again though, we'd need to trim the leading whitespace created.

## Summary

There's not much to say except use `unite()`!